---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
```{r setup, cache = F}
  library(knitr)
  opts_chunk$set(
    fig.align = "center")
```

```{r libraries}
  library(phoneticChange)
  library(plyr)
  library(dplyr)
  library(magrittr)
  library(ggplot2)
  library(grid)
  library(scales)
  library(RColorBrewer)
  library(shiny)

  # for gam smooths
  library(mgcv)
```

# Introduction

## The PNC

```{r fig.width = 8/1.5, fig.height = 5/1.5, out.width="90%"}
  ggplot(phoneticChange::white, aes(year, age, color = sex))+
    geom_point(position = position_jitter(width = 0.4))+
    theme_bw()
```


# /ay/ Raising

## General pattern

```{r calculate_ay_means}
  ays_means <- ays %>%
                filter(context=="internal", stress == 1) %>%
                group_by(idstring, year, age, DOB, sex, years_of_schooling, 
                         plt_vclass, word) %>%
                summarise(F1_n = mean(F1_n)) %>%
                summarise(F1_n = mean(F1_n))
```

```{r maxi_midlow_means}
  all_midlow_means <- mid_low_means%>%
                          group_by(plt_vclass)%>%
                          summarise(F1_n = mean(F1_n))%>%
                          mutate(label = c(uh="ʌ", o="\u0251")[plt_vclass])
          

```

```{r minimal_brewer_pal}
  three_col <- brewer.pal(3, "Dark2")
```

```{r fig.width = 8/1.25, fig.height = 5/1.25, out.width="90%", dev = 'CairoPNG'}
  ays_means %>% 
  ggplot(., aes(DOB, F1_n,
                      color = plt_vclass)) +
    #geom_point()+
    stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Normalized F1")+
    geom_hline(data = all_midlow_means, aes(yintercept = F1_n), 
               linetype = 3, color = "grey10")+
    geom_text(data  = all_midlow_means, x= 1895, aes(y = F1_n-0.1, label = label),
               color = "grey10")+
    scale_color_brewer("Vowel Category", palette = "Dark2")+
    xlab("Date of Birth")+
    theme_bw()
```


## What does it Mean

### Population Shift?

```{r}
  t1_pop <- data.frame(DOB = 1890, F1_n = rep(1.35,50), pos = "low")
  t2_pop <- data.frame(DOB = 1920, F1_n = c(rep(1.35,45), rep(0.35,5)),
                       pos =  c(rep("low",45), rep("high",5)))
  t3_pop <- data.frame(DOB = 1950, F1_n = c(rep(1.35,5), rep(0.35,45)),
                       pos =  c(rep("low",5), rep("high",45)))
  t4_pop <- data.frame(DOB = 1980, F1_n = c(rep(1.35,0), rep(0.35,50)),
                       pos =  c(rep("low",0), rep("high",50)))


  t_pop <- rbind_all(list(t1_pop, t2_pop, t3_pop, t4_pop))

```


```{r population, fig.width = 8/1.25, fig.height = 5/1.25, out.width="90%"}
  ays_means %>% 
    filter(plt_vclass == "ay0")%>%
  ggplot(., aes(DOB, F1_n)) +
    geom_point(data = t_pop, aes(color = pos), position = position_jitter(width = 2, height = 0.1))+
    stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Normalized F1")+
    scale_color_brewer("Speakers", palette = "Dark2", limits = c("low", "high"))+
    xlab("Date of Birth")+
    theme_bw()
```

## What would it look like?
### Population Shift

>- Speakers are not internally variable. You are either a raising speaker, or a non-raising speaker.
>- The speech community is highly variable. 

## What does it mean?
### Lexicon

```{r fifty_random_words}

  set.seed(5)
  ays %>%
    filter(plt_vclass == "ay0")%>%
    group_by(word)%>%
    summarise(n = n())%>%
    ungroup()%>%
    mutate(rand = rnorm(n()))%>%
    arrange(rand)%>%
    head(50)%>%
    mutate(word = tolower(word))%>%
    select(word)-> rand_fifty
```

```{r t_lexicon}
  t1_lex <- rand_fifty%>%
              mutate(DOB = 1890, F1_n =  rep(1.35,50), pos = "low")
              
  t2_lex <- rand_fifty%>%
              mutate(DOB = 1920, 
                     F1_n = c(rep(1.35,45), rep(0.35,5)),
                     pos =  c(rep("low",45), rep("high",5)))

  t3_lex <- rand_fifty%>%
              mutate(DOB = 1950, 
                     F1_n = c(rep(1.35,5), rep(0.35,45)),
                     pos =  c(rep("low",5), rep("high",45)))

  t4_lex <- rand_fifty%>%
              mutate(DOB = 1980, 
                     F1_n = c(rep(1.35,0), rep(0.35,50)),
                     pos =  c(rep("low",0), rep("high",50)))

  t_lex <- rbind_all(list(t1_lex, t2_lex, t3_lex, t4_lex))
```

```{r lexicon, fig.width = 8/1.25, fig.height = 5/1.25, out.width="90%"}
  ays_means %>% 
    filter(plt_vclass == "ay0")%>%
  ggplot(., aes(DOB, F1_n)) +
    geom_text(data = t_lex, 
              aes(color = pos, label = word), 
              position = position_jitter(width = 2, height = 0.1),
              size= 3)+
    stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Normalized F1")+
    scale_color_brewer("Lexicon", palette = "Dark2", limits = c("low", "high"))+
    xlab("Date of Birth")+
    theme_bw()
```

## What would it look like?
### Lexicon

>- Speakers are internally variable. Sometimes they'll say a raising word, and other times they'll say a non-raising word.
>- The speech communuty isn't very variable. Speakers should be more or less similar to eachother.

## What does it mean?
### Variable Rule

>- /ay/ = [+low]
>- with probability $p$, ay $\rightarrow$ [-low]/__[-voice]
    - beginning of the change, $p=0$
    - middle of the change, $p=0.5$
    - end of the change, $p=1$
    
    
## What does it mean?
### Variable Rule

```{r t_phon}
  t1_phon <- data.frame(DOB = 1890, F1_n = rep(1.35,50), pos = "[+low]")
  t2_phon <- data.frame(DOB = 1920, 
                        F1_n = c(rep(1.35,45), rep(0.35,5)),
                        pos =  c(rep("[+low]",45), rep("[-low]",5)))
  t3_phon <- data.frame(DOB = 1950, F1_n = c(rep(1.35,5), rep(0.35,45)),
                       pos =  c(rep("[+low]",5), rep("[-low]",45)))
  t4_phon <- data.frame(DOB = 1980, F1_n = c(rep(1.35,0), rep(0.35,50)),
                       pos =  c(rep("[+low]",0), rep("[-low]",50)))


  t_phon <- rbind_all(list(t1_phon, t2_phon, t3_phon, t4_phon))
```


```{r phon_var, fig.width = 8/1.25, fig.height = 5/1.25, out.width="90%"}
  ays_means %>% 
    filter(plt_vclass == "ay0")%>%
  ggplot(., aes(DOB, F1_n)) +
    geom_text(data = t_phon, 
              aes(color = pos, label = pos), 
              position = position_jitter(width = 2, height = 0.1),
              size= 3)+
    stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Normalized F1")+
    scale_color_brewer("Feature", palette = "Dark2", limits = c("[+low]", "[-low]"))+
    xlab("Date of Birth")+
    theme_bw()
```

## What would it look like?
### Variable Rule

>- Speakers would be internally variable, alternatively raising and not raising /ay/ with some internally consistant probability.
>- The speech communuty wouldn't be very variable. There would be some characteristic probability of raising for birth cohorts & social strata.

## What does it mean?
### Continuous Change

```{r setup_gam}
  ays_means %>% filter(plt_vclass == "ay0")%>%
    gam(F1_n ~ s(DOB, bs = "cs"), data = .)->gam_mod


  pred <- data.frame(DOB = rep(seq(1890, 1998, length = 10), 50))
  pred$F1_n <- predict(gam_mod, newdata = pred)

```

```{r continuous, fig.width = 8/1.25, fig.height = 5/1.25, out.width="90%"}
  ays_means %>% 
    filter(plt_vclass == "ay0")%>%
  ggplot(., aes(DOB, F1_n)) +
    geom_point(data = pred, 
              aes(color = F1_n), 
              position = position_jitter(width = 2, height = 0.1))+
    stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Normalized F1")+
    scale_color_gradient("Phonetic Target", high = three_col[1], low = three_col[2])+
    xlab("Date of Birth")+
    theme_bw()
```


## What would it look like?
### Continuous Change

>- Speakers wouldn't be very variable aside from the intrinsicw variation involved in trying to hit a real-valued target.
>- The speech community wouldn't very variable. There would be some characteristic pronunciation target for birth cohorts & social strata.

# What it does look like.

## Trend with speaker means

```{r trend_with_means, fig.width = 8/1.25, fig.height=5/1.25, out.width="90%"}
  ays_means %>% 
  ggplot(., aes(DOB, F1_n, color = plt_vclass)) +
    geom_point()+
    stat_smooth(method = "gam", formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Normalized F1")+
    scale_color_brewer("Vowel Class", palette = "Dark2")+
    xlab("Date of Birth")+
    theme_bw()
```

## Cohen's D

```{r}
  ays %>%
    filter(context == "internal")%>%
    group_by(idstring, DOB, plt_vclass)%>%
    summarise(mean = mean(F1_n), var = var(F1_n), n = n())%>%
    filter(n()==2)%>%
    arrange(plt_vclass)%>%
    summarise(s = sqrt(sum((n-1)*var)/(sum(n)-2)),
              diff = diff(mean))%>%
    mutate(d = diff/s)->ay_cohen

```

```{r cohen_plot, fig.width = 8/1.25, fig.height=5/1.25, out.width="90%"}
  ggplot(ay_cohen, aes(DOB, d))+
    geom_hline(y = 0)+
    geom_point()+
    stat_smooth(method = gam, formula = y ~ s(x, bs = "cs"))+
    scale_y_reverse("Cohen's d")+
    xlab("Date of Birth")+
    theme_bw()

```


## What it means

>- We can rule out the population shift model. 
>- The lexicon shift model, and the variable rule models can't be ruled out yet.
    - If speakers were sampling from two distributions, or two lexicons, their *means* would still shift continuously.
    
## Distributional Properties
